title:  The Ramer-Douglas-Peucker Algorithm
id:     5
date:   2014-03-03 22:00:00
tags:   algorithms, python
desc:   The Ramer-Douglas-Peucker algorithm visualized
javascripts: https://code.jquery.com/ui/1.10.4/jquery-ui.min.js,jquery.flot.js, rdp.js
stylesheets: https://code.jquery.com/ui/1.10.4/themes/smoothness/jquery-ui.min.css

A little while ago I was working on a project that dealt with the simplification
of rail tracks logged by GPS for the 
[Algorithm Engineering Group](http://www.algo.informatik.tu-darmstadt.de/algorithmics/?L=1)
of [Technische Universität Darmstadt](http://www.tu-darmstadt.de/index.en.jsp)
using data provided by [Deutsche Bahn](http://www.bahn.de).

Because the data contained too many coordinates, something had to be done in order
to reduce the number of points the rail tracks were representy by.
The Ramer-Douglas-Peucker algorithm
[@ramer1972iterative] and [@douglas1973algorithms],
an algorithm used for reducing the number of points in a curve,
came in quite handy for this task.

## Visualization

In the graph below you can see the algorithm in action. Move and release the slider in order
to see the results of the RDP algorithm applied to the problem at hand.

<div id="placeholder" style="width:600px;height:400px;align:middle;"></div>
<div id="sliderbox" style="width:600px"></div>
<div id="slider" style="width:570px;margin-left:15px;margin-top:10px;"></div>

<div style="margin-top:10px;margin-left:10px;">
<b>ε = </b><b id="epsilon"></b> <br />
<span style="color:#ba2762;font-weight:bold">Original # of points: </span><b id="c1"></b>
</br>
<span style="color:#659e5f;font-weight:bold">Simplified # of points: </span><b id="c2"></b>
</div>
<br />

## Python Implementation

I've created a [Python implementation of the RDP algorithm](https://pypi.python.org/pypi/rdp/),
which can be installed via pip:

    pip install rdp

And be used like this:

```{r engine="python"}
from rdp import rdp

res = rdp([[1, 1], [1, 1.1], [2, 2]], epsilon=0.5)
print(res)
```

Behind the scenes, this implementation uses [numpy](http://www.numpy.org/) and hence supports
passing a numpy array as well.

As far as the Point-Line-Distance in 2-Dimensional space is concerned, linear algebra comes
to the rescue. The implementation uses a custom distance function: If the line is specified
by two points $\textbf{x}_1 = (x_1, y_1)$ and $\textbf{x}_2 = (x_2, y_2)$, then the distance
from the point $\textbf{x}_0 = (x_0, y_0)$ is given by

$d = \frac{|\text{det}(\textbf{x}_2 - \textbf{x}_1 \textbf{x}_1 - \textbf{x}_0)|}{|\textbf{x}_2 - \textbf{x}_1|}$

where $\text{det}(A)$ denotes a determinant. The derivation of this formula can be found on 
[Wolfram MathWorld](http://mathworld.wolfram.com/Point-LineDistance2-Dimensional.html).
This can be translated into python as follows:
```{r engine="python"}
import numpy as np

def pldist(x0, x1, x2):
    if x1[0] == x2[0]:
        return np.abs(x0[0] - x1[0])

    return np.divide(np.linalg.norm(np.linalg.det([x2 - x1, x1 - x0])),
                     np.linalg.norm(x2 - x1))
```
